---
title: "MP2"
author: "Jonah Kotzen, Scarlett Varley, Ariana Roberts"
date: "2024-02-14"
output: html_document
bibliography: BIOL5380.bib
---

# Introduction

This project works to uncover how skeletal geometry and muscle length and mass affect both the force output in the forelimb and the maximum load it can support. The project aims to predict how much force and torque are produced by the biceps system over a range of extension/flexion angles, assess how force and torque vary according to extension/flexion angle, and test these predictions in the gym. The question will be answered through computational analysis of torque, force, and maximum mass at respective angles with the measurements of arm length, inlever length, outlever length, and weight of bicep. 

# Methods

Arm length, inlever length, and outlever length were measured in lab using a meter stick. Bicep weight was predicted by weighing play-dough and approximating the size of each participant's bicep muscle. Computational analysis is as follows usuing the Hill-type muscle model [@hill1938not]. Additionally, we utilized relationship approximation between force and muscle length for mammalian skeletal muscle [@whiting1999not]. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```


```{r load, include = FALSE}

arm_dat <- read_csv("MP2.csv")

# Convert cm to m and g to kg
arm_dat <- arm_dat %>%
  mutate(
    Larm = Larm / 100,
    Li = Li / 100,
    Lout = Lout/ 100,
    Mmuscle = Mmuscle / 1000
  )

``` 

### Dataset for 3 Subjects (Jonah, Scarlett, Ari)

```{r dataset, include=TRUE}

print(arm_dat)

```

### Functions for Calculations

```{r chunk 1, include=TRUE}

law_cos <- function(l, r, o) {
  theta <- acos((-o^2 + l^2 + r^2) / (2 * l * r))
  return(theta)
}

law_cos2 <- function(l, r, theta) {
  o <- sqrt(l^2 + r^2 - 2 * l * r * cos(theta))
  return(o)
}

force_length_factor <- function(Lmuscle, Lmuscler) {
  Lopt <- Lmuscler * 0.8
  ffl <- -6.25 * (Lmuscle / Lopt)^2 + 12.5 * (Lmuscle / Lopt) - 5.25
  return(ffl)
}

calculate_Fmax <- function(Mmuscle, Lmuscler, rho = 0.00105, phi = 0, k = 80) {
  PCSA <- (Mmuscle * cos(phi)) / (rho * Lmuscler)
  Fmax <- PCSA * k
  return(Fmax)
}


```


### The Model Function

```{r chunk5, include=TRUE}
model <- function(Mmuscle, Lmuscler, Lmuscle, theta, Li, rho = 0.00105, k = 80) {
   # Convert Lmuscler from m to cm for PCSA calculation (as ρ is given in kg/cm^3)
  Lmuscler_cm <- Lmuscler * 100
  # Assuming phi is 0 as the muscle is nonpennate
  phi <- 0
  # PCSA calculation
  PCSA <- (Mmuscle * cos(phi)) / (rho * Lmuscler_cm)
  # Fmax calculation
  Fmax <- PCSA * k
  # Force-length relationship factor (ffl)
  ffl <- force_length_factor(Lmuscle, Lmuscler)
  # Muscle force (Fmuscle)
  Fmuscle <- Fmax * ffl
  # Torque calculation (τmuscle)
  tau_muscle <- Fmuscle * sin(theta) * Li
  return(tau_muscle)
}
```


# Results

The results obtained display the length of bicep muscle, muscle torque, max mass for degree angles 50, 75, 125, and 150 for each person. The two graphs display muscle torque (Nm) at each flexion angle and max mass (kg) at each flexion angle. 

### Final Tibble

```{r chunk6, include=FALSE}
angles_tibble <- tibble(
  flexion_deg = c(50, 75, 100, 125, 150),
  flexion_rad = flexion_deg * pi / 180
)

# Assuming g = 9.81 m/s^2
g <- 9.81

# Calculate Lmuscle, theta, muscle_torque, and max_mass for each flexion angle
dat1 <- arm_dat %>%
  cross_join(angles_tibble) %>%
  rowwise() %>%
  mutate(
    Lmuscle = law_cos2(Li, Larm, flexion_rad), 
    theta = law_cos(Li, Lmuscle, Larm)) %>% 
  ungroup()
  
dat2 <- dat1 %>% 
filter (flexion_deg == 150) %>%
mutate(Lmuscler = Lmuscle) %>%
select (subject, Lmuscler)

dat3 <- dat1 %>%
  left_join(dat2, by = "subject")

dat_final <- dat3 %>%
  mutate(
    muscle_torque = model(Mmuscle, Lmuscler, Lmuscle, theta, Li),
    max_mass = muscle_torque / (Lout * g)
  ) %>%
  ungroup()

dat <- dat_final %>%
  select(subject, flexion_rad, Lmuscle, theta, muscle_torque, max_mass)

```

```{r final tib, include=TRUE}

print(dat)

```

### Graphs

```{r vis, include=TRUE}

# Plot muscle torque for each flexion angle for each subject
ggplot(dat_final, aes(x = flexion_deg, y = muscle_torque, group = subject, color = subject)) +
  geom_line() +
  geom_point() +
  labs(title = "Muscle Torque at Each Flexion Angle",
       x = "Flexion Angle (degrees)",
       y = "Muscle Torque (Nm)")

# Plot max mass for each flexion angle for each subject
ggplot(dat_final, aes(x = flexion_deg, y = max_mass, group = subject, color = subject)) +
  geom_line() +
  geom_point() +
  labs(title = "Max Mass at Each Flexion Angle",
       x = "Flexion Angle (degrees)",
       y = "Max Mass (kg)")


```

# Discussion 

### What were your predicted τ muscle and Mload values at each flexion angle?
- The predicted muscle torque and Mload values at each flexion angle are displayed in both the table and graphs above. 


### Is there any pattern to these values? What explains these patterns?

- The data shows a trend where muscle torque and max mass generally peak around 75 degrees of flexion for all three subjects, indicating an optimal angle for force generation due to favorable muscle length-tension relationships and lever mechanics. Beyond this angle, both torque and mass support capability decrease, likely due to reduced leverage and less effective muscle contraction angles. This pattern highlights the significance of biomechanics in understanding physical performance.


### How did your gym testing compare to your model estimations? That is, was it more or less than predicted?

- In gym testing, we were able to lift more than the model predicted. 


### If the model was produced poor predictions, why?

- This discrepancy could be due to the model not accounting for synergistic muscle actions, neural adaptations, individual biomechanical differences, and psychological factors that can enhance performance. Essentially, real-world performance can exceed model predictions due to the complex interplay of physiological and psychological factors not fully captured by the model. Additionally, other muscles present in the upper arm and shoulder could have compensated for the bicep and aided in the lifting, resulting in participants being able to lift a heavier mass than predicted. 


# Author Contributions


# References


