---
title: "MP2"
author: "Jonah Kotzen, Scarlett Varley, Ariana Roberts"
date: "2024-02-14"
output: html_document
bibliography: BIOL5380.bib
---

# Introduction

This project works to uncover how skeletal geometry and muscle length and mass affect both the force output in the forelimb and the maximum load it can support. The project aims to predict how much force and torque are produced by the biceps system over a range of extension/flexion angles, assess how force and torque vary according to extension/flexion angle, and test these predictions in the gym. The question will be answered through computational analysis of torque, force, and maximum mass at respective angles with the measurements of arm length, inlever length, outlever length, and weight of bicep. 

# Methods

Arm length, inlever length, and outlever length were measured in lab using a meter stick. Bicep weight was predicted by weighing play-dough and approximating the size of each participant's bicep muscle. Computational analysis is as follows usuing the Hill-type muscle model [@hill1938heat]. Additionally, we utilized the relationship approximation between force and muscle length for mammalian skeletal muscle [@whiting1999theoretical]. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```


```{r load, include = FALSE}

arm_dat <- read_csv("MP2.csv")

# Convert cm to m and g to kg
arm_dat <- arm_dat %>%
  mutate(
    Larm = Larm / 100,
    Li = Li / 100,
    Lout = Lout/ 100,
    Mmuscle = Mmuscle / 1000
  )

``` 

### Dataset for 3 Subjects (Jonah, Scarlett, Ari)

```{r dataset, include=TRUE}

print(arm_dat)

```

### Functions for Calculations

```{r chunk 1, include=TRUE}

law_cos <- function(l, r, o) {
  theta <- acos((-o^2 + l^2 + r^2) / (2 * l * r))
  return(theta)
}

law_cos2 <- function(l, r, theta) {
  o <- sqrt(l^2 + r^2 - 2 * l * r * cos(theta))
  return(o)
}

force_length_factor <- function(Lmuscle, Lmuscler) {
  Lopt <- Lmuscler * 0.8
  ffl <- -6.25 * (Lmuscle / Lopt)^2 + 12.5 * (Lmuscle / Lopt) - 5.25
  return(ffl)
}

calculate_Fmax <- function(Mmuscle, Lmuscler, rho = 0.00105, phi = 0, k = 80) {
  PCSA <- (Mmuscle * cos(phi)) / (rho * Lmuscler)
  Fmax <- PCSA * k
  return(Fmax)
}


```


### The Model Function

```{r chunk5, include=TRUE}
model <- function(Mmuscle, Lmuscler, Lmuscle, theta, Li, rho = 0.00105, k = 80) {
   # Convert Lmuscler from m to cm for PCSA calculation (as ρ is given in kg/cm^3)
  Lmuscler_cm <- Lmuscler * 100
  # Assuming phi is 0 as the muscle is nonpennate
  phi <- 0
  # PCSA calculation
  PCSA <- (Mmuscle * cos(phi)) / (rho * Lmuscler_cm)
  # Fmax calculation
  Fmax <- PCSA * k
  # Force-length relationship factor (ffl)
  ffl <- force_length_factor(Lmuscle, Lmuscler)
  # Muscle force (Fmuscle)
  Fmuscle <- Fmax * ffl
  # Torque calculation (τmuscle)
  tau_muscle <- Fmuscle * sin(theta) * Li
  return(tau_muscle)
}
```


# Results

The results obtained display the length of bicep muscle, muscle torque, max mass for degree angles 50, 75, 125, and 150 for each person. The two graphs display muscle torque (Nm) at each flexion angle and max mass (kg) at each flexion angle. 

### Final Tibble

```{r chunk6, include=FALSE}
angles_tibble <- tibble(
  flexion_deg = c(50, 75, 100, 125, 150),
  flexion_rad = flexion_deg * pi / 180
)

# Assuming g = 9.81 m/s^2
g <- 9.81

# Calculate Lmuscle, theta, muscle_torque, and max_mass for each flexion angle
dat1 <- arm_dat %>%
  cross_join(angles_tibble) %>%
  rowwise() %>%
  mutate(
    Lmuscle = law_cos2(Li, Larm, flexion_rad), 
    theta = law_cos(Li, Lmuscle, Larm)) %>% 
  ungroup()
  
dat2 <- dat1 %>% 
filter (flexion_deg == 150) %>%
mutate(Lmuscler = Lmuscle) %>%
select (subject, Lmuscler)

dat3 <- dat1 %>%
  left_join(dat2, by = "subject")

dat_final <- dat3 %>%
  mutate(
    muscle_torque = model(Mmuscle, Lmuscler, Lmuscle, theta, Li),
    max_mass = muscle_torque / (Lout * g)
  ) %>%
  ungroup()

dat <- dat_final %>%
  select(subject, flexion_rad, Lmuscle, theta, muscle_torque, max_mass)

```

```{r final tib, include=TRUE}

print(dat)

```

### Graphs

```{r vis, include=TRUE}

# Plot muscle torque for each flexion angle for each subject
ggplot(dat_final, aes(x = flexion_deg, y = muscle_torque, group = subject, color = subject)) +
  geom_line() +
  geom_point() +
  labs(title = "Muscle Torque at Each Flexion Angle",
       x = "Flexion Angle (degrees)",
       y = "Muscle Torque (Nm)")

# Plot max mass for each flexion angle for each subject
ggplot(dat_final, aes(x = flexion_deg, y = max_mass, group = subject, color = subject)) +
  geom_line() +
  geom_point() +
  labs(title = "Max Mass at Each Flexion Angle",
       x = "Flexion Angle (degrees)",
       y = "Max Mass (kg)")


```

# Discussion 

The predicted muscle torque and Mload values at each flexion angle are displayed in both the table and graphs above. The data shows a trend where muscle torque and max mass generally peak around 75 degrees of flexion for all three subjects, indicating an optimal angle for force generation due to favorable muscle length-tension relationships and lever mechanics of the bicep muscle. This angle for the bicep allows for maximum overlap of myosin filaments along with actin that are important for force and muscle contraction and production. When exceeding 75 degrees, both torque and mass support capability decrease, likely due to reduced leverage as the forearm length decreases also decreasing the mechanical advantage of the bicep which is why it is more difficult ot hold or lift heavier weights at that position. Tension in the bicep also decreases as the muscle fibers are not being fully engaged which limits the maximum mass as well confirming the significance of biomechanics in understanding physical performance which can also help understand the proper engagement to lifting weights when working out. Understanding the composition of the bicep, along with factors such as angle, torque, and load, is essential for determining the proper flexion angle.

In gym testing, there was a correlation that the females were on average able to lift more than the model predicted. Ari was able to lift approximately 2.3 kg above the calculated maximum mass, and Scarlett was able to lift approximately 2.5 kg above the calculated maximum mass. Then for males or the other group mate Jonah, he lifted approximately 2.7 kg under the predicted max mass. This could be due to over approximation of bicep mass in addition to inaccurate conversion when at the gym. Furthermore, individual differences in muscle composition, arm lengths, and can even have different anatomy of surrounding muscles, also can be influenced by varying levels of bicep workouts, may also contribute to differences in lifting capacity. Scarlett and Ari also tested the predicted model together at the gym, Jonah collected his data independently, potentially leading to variations that could have occurred during data collection. The Play-Doh seemed to do some sort of justice, but was definitely hard to predict the correct mass which is extremely influential to our model.

To elaborate, the discrepancy could be due to the model not accounting for synergistic muscle actions, neural adaptations, individual biomechanical differences, and psychological factors that can enhance performance. Essentially, real-world performance can exceed model predictions due to the complex interplay of physiological and psychological factors not fully captured by the model. Additionally, as far as Ari and Scarlett, other muscles present in the upper arm and shoulder could have compensated for the bicep and aided in the lifting, resulting in participants being able to lift on average a heavier mass than predicted. It could have been less for Jonah as problems could have arisen in using the opposite arm than the one used for predicted measurements as most of the time one side of you body is more dominant than the other, it also could have been that Jonah measured differently than from how Scarlett and Ari measured as they measured together for the predicted values. It is extremely difficult to predict the measurements when each person measures differently and accounts for every aspect of the measurements differently, something to consider would be getting a more accurate and precise method to make these predictions, but from our data it seems justifiable.  

Further into that many advancements have been made in the the analysis of musculoskeletal systems, including the Hill-type muscle model studied by @hill1938heat and the force and muscle length approximation equation studied by @whiting1999theoretical. Hill emphasizes the critical role of technique, which must be consistent across all individuals using this model. Technique is a complex factor that significantly influences the accuracy of results, especially when considering the viscoelastic properties and deeper muscle anatomy. While our general model lacks specificity, it adequately fulfilled its purpose. However, our reliance on approximate weights provided by the gym, which necessitated rounding up or down, could potentially lead to inaccuracies in the results. Whiting's equations would have been useful to ensure that everyone measured in a consistent manner, which could have led to more consistent predicted measurements. However, since we all measured in our own way, this variability may have skewed the data. So the data supports how more research must be done in order to better analyze and predict how skeletal geometry and muscle length and mass affect both the force output in the forelimb and the maximum load it can support. 


# Author Contributions


# References

